Clinical prediction using similarity-based cohorts: A scoping review

### 2023-08-13 docs/Manuscript/Manuscript.Rmd
begin making edits in response to reviewers

We discuss in this review a hybrid computational approach we call _localized modeling_ that borrows elements from both clinical application areas and appears to have been introduced independently several times. The approach fits ML models, including regularized families of generalized regression models (GLMs), to cohorts retrieved via similarity matching. The idea is to use patient similarity not to obtain an ideally homogeneous population matched to the index case, as in CBR and NN prediction, but to obtain a relevant yet heterogeneous population, each member of which deviates slightly from the index case. As a result, the variation of any single factor will be (1) centered near its value on the index case and (2) most different from the index case in cases that in other respects are more similar to it. This makes interpretative models like GLMs better able to identify and measure the effects of determining factors. Localized modeling shows promise toward simultaneously achieving the performance improvements of ML and the interpretability of classical models.

One memorable aspiration of clinical informatics was the "green button" that would, in response to clinical queries, automatically retrieve patient data from an institution’s records with which to conduct on-demand retrospective studies for an individual patient @Longhurst2014. Localized modeling has recently been put forth as one answer to this appeal. Studies using localized models have been limited by computational cost—a naïve implementation involves fitting, optimizing, and evaluating a model on a separate population for each training case—and recent studies have described algorithms and implementations designed to reduce this cost. As greater access to health data, informatical knowledge, algorithmic efficiency, and computational power continue to spur innovation in CDS, we may expect localized modeling to become more accessible and efficient, its advantages better understood, and its disadvantages ameliorated.

The motivation for this scoping review is to describe the settings in and problems with which localized models have been tasked to date, and to synthesize existing methods into a general framework. Along the way, we attempt to reconcile varying terminology, summarize evaluation techniques, propose needed follow-up work, and recommend best practices.

>>>>

We discuss in this review a hybrid computational approach we call _localized modeling_ that borrows elements from both clinical application areas and appears to have been introduced independently several times. The approach fits ML models, including regularized families of generalized regression models (GLMs), to cohorts retrieved via similarity matching. Whereas CBR and NN prediction use patient similarity to obtain an ideally homogeneous subpopulation matched to the index case, in localized modeling one seeks a relevant yet heterogeneous subpopulation, each member of which deviates slightly from the index case. In principle, this allows interpretable models like GLMs to identify and measure the effects of determining factors that are specific to the index case. In the reviewed studies, localized modeling shows some promise toward simultaneously achieving the performance improvements of ML and the interpretability of classical models.

One memorable aspiration of clinical informatics was the "green button" that would, in response to a query, automatically retrieve patient data from an institution’s records with which to conduct on-demand retrospective studies for an individual patient @Longhurst2014. Localized modeling has recently been put forth as one answer to this appeal. Studies using localized models have been limited by computational cost—a naïve implementation involves fitting, optimizing, and evaluating a model on a separate population for each training case—and recent studies have described algorithms and implementations designed to reduce this cost. As greater access to health data, informatical knowledge, algorithmic efficiency, and computational power continue to spur innovation in CDS, we may expect localized modeling to become more accessible and efficient, its advantages better understood, and its disadvantages ameliorated.

The motivation for this scoping review is to describe the settings in and problems with which localized models have been tasked to date, and to synthesize existing methods into a general framework. Along the way, we attempt to reconcile terminology, summarize evaluation techniques, propose needed follow-up work, and recommend best practices.
